Subgroup detection in linear growth curve models with generalized linear mixed model (GLMM) trees

### 2023-09-08 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
LGCM paper updates
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3276 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: bb13f5e03206caef93d0ccd9295ce5e69cdf2d4a

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}). 
Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability. For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept. For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept. Also, the two groups (or nodes) with higher intercepts have higher slopes, indicating that children with relatively higher ability may also show larger gains in ability over time.

@@

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals.}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
<<fig=TRUE, height=8, width=14, echo=FALSE>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

@@

In the next sections, we will assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.

>>>>

Recursive partitioning methods (RPMs), also known as ``trees``, allow for identifying relevant predictors from a potentially (very) large number of covariates. Figure~\ref{fig:LMM_tree_r} shows an example tree, which identified socio-economic status (SES), gross motor skills (GMOTOR) and internalizing problems (INTERN) from a set of 11 socio-demographic and behavioral characteristics of the children, assessed at baseline. Five subgroups were identified, corresponding to the terminal nodes of the tree, each with a different estimate of the fixed intercept and slope %(presented in Table~\ref{tab:local_coefs}). 
Groups of children with lower (higher) SES also have a lower (higher) intercept, indicating lower (higher) average science ability. For children with lower SES (node~2), there is also an effect of gross motor skills: The group with lower motor skills has a lower intercept. For children with intermediate levels of SES (node~6), there is also an effect of internalizing problems: The group with lower internalizing problems has a higher intercept. The two groups (or nodes) with higher intercepts also have higher slopes, indicating that children with higher ability also gain more ability over time.

@@

## Plot tree
plot(lmm_tree_r, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1.5))
@
\label{subfig:tree}
\end{subfigure}%
~
\begin{subfigure}{0.25\textwidth}%
\vspace{-2.95cm}
\hspace{-2.5cm}
\scriptsize
<<echo=FALSE, results=tex, fig.pos='b'>>=
library("kableExtra")
coefs <- coef(lmm_tree_r)
coefs <- data.frame(cbind(Node = as.numeric(rownames(coefs)), coefs))
colnames(coefs) <- c("Node", "Intercept", "Slope")
coefs$Slope <- paste0("$", formatC(coefs$Slope, digits = 3, format = "f"), "$")
coefs$Intercept <- paste0("$", formatC(coefs$Intercept, digits = 3, format = "f"), "$")
coefs$Intercept[!grepl("-", coefs$Intercept)] <- paste("$\\:\\:$",
                                                       coefs$Intercept[!grepl("-", coefs$Intercept)])
kable(coefs, format = "latex", linesep = "", # linesep command suppressess addlinesep every 5 rows
        row.names = FALSE, digits = 3L,escape=FALSE, align = c("ccc"), booktabs = TRUE)
@
\label{subfig:coefs}
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, $y$-axes represent science ability. Gray lines depict observed individual trajectories, red lines depict average growth curve within each terminal node, as estimated with a linear mixed-effect model comprising node-specific fixed effects of time and a random intercept with respect to individuals. The table presents numerical estimates of fixed intercepts and slopes.}
\label{fig:LMM_tree_r}
\end{figure}%

@@

\begin{figure}%
\caption{GLMM tree estimated by initializing with zero random effects.}
\begin{subfigure}{1.2\textwidth}
<<fig=TRUE, height=9, width=16, echo=FALSE>>=
lmm_tree <- lmertree(LMM_form, data = data, maxdepth = 6)
plot(lmm_tree, which = "growth", ip_args = list(pval = FALSE), 
     nodesize_level = "CHILDID", tp_args = 
       list(xscale = c(0, 60^(2/3)), xaxis.at = c(0, 12, 36, 60)^(2/3),
            xaxis.labs = c(0, 12, 36, 60)), gp = gpar(cex = 1))
icc <- as.data.frame(VarCorr(lmm_tree))
lmm_preds <- predict(lmm_tree, newdata = testdat, re.form=NA)
@
\end{subfigure}
\\\footnotesize{\textit{Note.} The $x$-axes represent the number of months after the baseline assessment, the $y$-axes represent science ability. Gray lines depict observed individual trajectories. Red lines depict the average growth curve within each terminal node, as estimated with a linear mixed-effect model with a node-specific fixed effect of time and random intercepts estimated with respect to individuals.}
\label{fig:LMM_tree}
\end{figure}%

@@

In the next sections, we assess performance of the original algorithm and proposed adaptations through more extensive empirical evaluations.
